---
title: "01_load"
format: html
editor: visual
---

# Load raw data from the "Genes" dataset:
The original data is structured in the following manner:
- The "Genes" folder contains a number of excel files.
- Each excel file corresponds to a persistant organic pollutant (POP) compound.
- Each excel file contains a number of sheets.
- Each excel sheet corresponds to a gene targeting bioremediation of said POP.
- Each excel sheet contains 28 variables and a number of observations.
```{r}
# We create a function that reads all sheets from a given excel file:
read_excel_sheets <- function(file_path) {
  sheets <- excel_sheets(file_path)
  map(sheets,
      ~ read_excel(file_path, 
                           sheet = .x))
}

# Now, we make a list of all files inside the "Genes" folder:
folder_path <- "../data/_raw/Genes" 
excel_files <- list.files(folder_path, 
                          pattern = "\\.xlsx$", 
                          full.names = TRUE)

# Finally, we apply the read_excel_sheets function to every excel file in the excel_files list:
all_sheets <- map(excel_files, 
                  read_excel_sheets)

```
The final object is a list of lists of excel sheets data. The outer list contains filenames, the inner list contains file_sheets, and each sheet contains the variables and data.


# Join data into a single table:
```{r}
# Initialize the joined_data with the first sheet
joined_data <- all_sheets[[1]][[1]]

# Iterate through the remaining sheets and join them to the joined_data
joined_data <- reduce(all_sheets[-1], ~ full_join(.x, .y))
```

