---
title: "Sequence analysis"
format:
  html:
    embed-resources: true
editor: visual
---

# Sequence analysis

## Load libraries:

```{r}
library("tidyverse")
library("msa")
library("ape")
library("seqinr")
library("ggseqlogo")
source("99_proj_func.R")
```

## Load augmented dataset from step 4

```{r}
tsv_file <- "../data/data_augmented.tsv"
data_augmented <- read_tsv(tsv_file)
```

## Group sequences by compound

Create a new dataset where the sequences from individual samples are joined in the same group if the organism degrade the same compound. Filter samples without sequence.

```{r}
compound_lists <- data_augmented |>
  filter(!is.na(Sequence)) |> 
  group_by(Compound_Name) |> 
  summarise(ProteinSequences = list(Sequence))
```

## Apply MSA on each compound

```{r}
# Perform MSA on each list of sequences
aligned_compounds <- compound_lists |>
  filter(map_lgl(ProteinSequences, ~length(.x) < 50)) |> 
  mutate(MSA = map(ProteinSequences, ~msa(.x, type = "protein")))
```

## Consensus sequence

Extract consensus sequence from MSA and check the percentage of gaps and missing positions.

```{r}

aligned_compounds <- aligned_compounds |>
  mutate(consensus_sequence = map(MSA, ~consensusString(consensusMatrix(.x)))) |>
  mutate(percentage = unlist(map(consensus_sequence, ~calculate_percentage(.x))))
  
```

Plot the percentages for each compound

```{r}
aligned_compounds |>
  ggplot(aes(x = Compound_Name, y = percentage)) + 
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1, vjust = 1), axis.text.y = element_text(size = 8))
    labs(title = "Percentage of missing positions for each compound alignment", x = "Compounds", y = "Percentage")
```

## Aligned sequences

We extract aligned sequences from MSA alignment to perform posterior analysis

```{r}
aligned_compounds <- aligned_compounds |>
  mutate(aligned_sequences = map(MSA, ~msaConvert(.x, type="bios2mds::align"))) |>
  mutate(aligned_sequences = map(aligned_sequences, ~map(.x, ~paste(.x, collapse = ""))))
```

## Distance matrix

```{r}
aligned_compounds <- aligned_compounds |>
  mutate(distance_matrix = map(MSA, ~msaConvert(.x, type="seqinr::alignment"))) |>
  mutate(distance_matrix = map(distance_matrix, ~dist.alignment(.x, "identity")))
```

## Phylogenetic trees

```{r}

phylo_tree <- njs(aligned_compounds$distance_matrix[6][[1]])

# Plot the tree
plot(phylo_tree, main="Phylogenetic Tree of a compound example", cex=0.5)

```


## Core sequences

In order to make sequence logos on the core sequences, we need to calculate the best 9mer.

```{r}
aligned_compounds <- aligned_compounds |>
  mutate(conMat = map(MSA, ~consensusMatrix(.x))) |>
  mutate(mer_position = map(conMat, ~mer_position_function(.x)))
```

```{r}
aligned_compounds <- aligned_compounds |>
  mutate(core = map2(aligned_sequences, mer_position, ~map2(.x, rep(.y, length(.x)), ~str_sub(., .y, .y + 8)))) |>
  mutate(core = map(core, ~unlist(.x)))

aligned_compounds <- aligned_compounds |>
  mutate(seqlogo = map(core, ~ggseqlogo(.x, seq_type = "aa")))
```

